Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-26 _paper_output/paper_sparse_hawkes.tex
make everything nice and explained

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -\/-\/-
\cite{ozaki_maximum_1979}. Here, however, we use automatic
differentiation to make more complex penalisation procedure convenient
to differentiate \cite{baydin_automatic_2014}. The optimisation proceeds
using a standard black-box based Truncated Newton algorithm; as more
specialised sparse regression algorithms such as Least-Angle Regression,
do not directly apply here.

@@

The degrees of freedom for parametric models is taken to be the number
of free parameters, equal to the cardinality of the parameter vector
\(\theta\).

@@

We proceed heuristically, based on the degrees-of-freedom arguments from
Lasso-type linear-model regularisation. There, the number of non-zero
coefficients fit under a particular value of the regularization penalty
\(\pi\) in an \(\ell_1\) regression model provides an unbiassed estimate
of the effective degrees of freedom of the model
\cite{zou_degrees_2007}.

@@

\[
\hat{d}\left(N,\hat{\theta},\pi\right)=\sum_j\mathbb 1\left\{\hat{\theta_j}(N,p,q)\neq 0\right\}
\]

@@

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

@@

    \begin{Verbatim}[commandchars=\\\{\}]

        ---------------------------------------------------------------------------

        FileNotFoundError                         Traceback (most recent call last)

        <ipython-input-8-765ef5deb338> in <module>()
          1 from skopt import load
          2 
    ----> 3 res\_gp = load(join(FIT\_DIR, 'additive\_cvge.gp.bz2'))

        /usr/local/lib/python3.6/site-packages/skopt/utils.py in load(filename, **kwargs)
        122         Reconstructed OptimizeResult instance.
        123     """
    --> 124     return load\_(filename, **kwargs)

        /usr/local/lib/python3.6/site-packages/sklearn/externals/joblib/numpy\_pickle.py in load(filename, mmap\_mode)
        565             obj = \_unpickle(fobj)
        566     else:
    --> 567         with open(filename, 'rb') as f:
        568             with \_read\_fileobject(f, filename, mmap\_mode) as fobj:
        569                 if isinstance(fobj, \_basestring):

        FileNotFoundError: [Errno 2] No such file or directory: '/Users/dan/Datasets/branching\_process\_output/additive\_cvge.gp.bz2'

    \end{Verbatim}

>>>>

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -\/-\/-
\cite{ozaki_maximum_1979}. Here, however, we use standard reverse-mode
automatic differentiation \cite{baydin_automatic_2014}. The optimisation
proceeds using a standard gradient-based black-box based Truncated
Newton algorithm; as more specialised sparse regression algorithms such
as Least-Angle Regression, do not directly apply here.

@@

The degrees of freedom for a parametric model is taken to be the number
of free parameters, equal to the cardinality of the parameter vector
\(\theta\).

@@

We proceed heuristically, based on the degrees-of-freedom arguments from
Lasso-type linear-model regularisation. There, the number of non-zero
coefficients fit under a particular value of the regularization penalty
\(\pi\) in an \(\ell_1\) regression model provides an unbiased estimate
of the effective degrees of freedom of the model
\cite{zou_degrees_2007}.

@@

\[
\hat{d}\left(N,\hat{\theta},p,q\right)=\sum_j\mathbb 1\left\{\hat{\theta_j}(N,p,q)\neq 0\right\}
\]

@@

    TODO: switch to consistent step-function display of instantaneous rate.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_23_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\texttt{\color{outcolor}Out[{\color{outcolor}27}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_25_0.png}
    \end{center}
    { \hspace*{\fill} \\}

    With a single regularisation parameter, only for the background rate
coefficient \(\omega_j\), we recover the background rate efficiently.

\texttt{\color{outcolor}Out[{\color{outcolor}28}]:}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_27_0.png}
    \end{center}
    { \hspace*{\fill} \\}

    \subsection{How to simulate a Hawkes
process}\label{how-to-simulate-a-hawkes-process}

Here we follow a standard "thinning" method of simulating Hawkes
processes, given as Algorithm 7.5.IV of Daley and Vere-Jones'
\emph{Introduction to the Theory of Point Processes Volume 1}
\cite{daley_introduction_2003}, restating Ogata's original Algorithm 2
(\cite{ogata_lewis_1981}.

Since that is a general case, here we present a specific version for
linear Hawkes processes which is considerably simpler.

For clarity, we will use the representation of the Hawkes process as an
immigrant-birth process, where the recursive "self-excited" component
corresponds to the "births". We will assume that we are given a list of
arrival times of immigrants, \(\{a_1,a_2,\dots,a_N\}\), and wish to
simulate the births arising from these immigrants. We will associate a
point and its arrival time and discuss them interchangeably where
ambiguity does not arise.

Each of these \(a_i\) gives rise to a family of descendants
\(\mathrm{descendants}(a_i)\), distributed independently of all other
families, whose combined superposition is the final Hawkes process
(\cite{hawkes_cluster_1974}. It will suffice to simulate the arrival
times of each of these families. Without loss of generality we may
consider the problem of simulating the births in a single family started
at time \(a_1=0\) and translate it.

Now the families
\(\mathrm{descendants}(t_j)\forall t_j\in \mathrm{children}(a_i)\) are
all once again identically and mutually independently distributed.
Proceeding by induction, it suffices to simulate the children of a
single point at time \(0\). The same method will suffice for its
children, children of children and so on. The final realisation can be
constructed from these simualtions by appropriate concatenation and
translation of the generations of each family.

The single kernel simulation is simple, since the Poisson intensity
process is no longer conditional on the history of the sequence. As
before, the influence kernel is \(\phi\). We will assume that \(\phi\)
is a.e. continuous \(L_1\) integrable and eventually decreasing in
\(t\). We associate with \(\phi\) a "majorant", \(\phi^{**}\) such that
\[\phi^{**}(t)\geq\phi(t), \forall t>0\] and \(\phi^{**}\) is
non-increasing and integrable. With the unimodal influence kernels used
in the current paper, such majorants are trivial to construct.

We will use the fact that there is some time \(T>0\) such that we do not
consider events after it, i.e. that we only care about arrival times on
\([0,T]\), and so only children that occur before time \(T\) can be
relevent.

\subsubsection{Algorithm: Ogata's thinning
algorithm}\label{algorithm-ogatas-thinning-algorithm}

Our goal is to simulate the set \(S=\mathrm{children}(a_0)\).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set \(i=1, t=0\), \(S=\varnothing\)
\item
  Generate \(\Delta\sim \operatorname{Exp}(1/\phi^{**}(t))\)
\item
  Set \(t\leftarrow t+\Delta\)
\item
  If \(t\geq T\), halt.
\item
  With probability \(P=\frac{\phi(t)}{\phi^{**}(t)}\), set
  \(S\leftarrow S\cup \{t\}\).
\item
  Continue from step 2.
\end{enumerate}

@@
